---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
#read in data file
  data1 <- read.csv("./activity.csv")
```


## What is mean total number of steps taken per day?
```{r}
#Summarize Steps by Date to a dataset called 't'
t <- aggregate(steps ~ date, data=data1, FUN=sum)

#Create histogram
#Histogram of steps taken
hist(t$steps,xlab = "Total Steps", main = "Total steps per day")
```

Calculate the mean and median steps taken per day
```{r}
##Summary Stats steps
mean(t$steps)
median(t$steps)
```



## What is the average daily activity pattern?
```{r}
#Summarize Steps by Date & Interval
t2 <- aggregate(steps ~ interval, data=data1, FUN=sum)

#count number of days in data source
uniq<- length(unique(data1$date))
#61 days

#create new variable with average steps
t2$avgsteps <- t2$steps/uniq


plot(t2$interval, t2$avgsteps, type="l",
     ylab="Avg Steps", xlab="time",
     main = "Steps in 5 Minute Increments")

```


Calculate time increment with maximum avg steps

```{r}
t2[which.max(t2$avgsteps),] 
#835
```

The maximum occurs at interval 835

## Imputing missing values

Determine number of missing values
```{r}
summary(data1)
```

2,304 records with NA

####Impute missing data (NA) with the median values of steps at that specific interval
```{r}
#compute median step time for imputing
step_med <- aggregate(steps ~ interval, data=data1, FUN=median)

#change colnames
colnames(step_med) <- c("interval","med_steps")


library("plyr")
library("dplyr")

#Append median time by merging based on interval
Joindata<-arrange(join(data1,step_med),interval)


Joindata$steps[is.na(Joindata$steps)] <- Joindata$med_steps[is.na(Joindata$steps)]
#remove column with median steps for imputing
Joindata$med_steps <- NULL

#Summarize Steps by Date
Joindata_t <- aggregate(steps ~ date, data=Joindata, FUN=sum)
```
####Create histogram
```{r}
##Histogram of steps taken
hist(Joindata_t$steps,xlab = "Total Steps", main = "Total steps per day (with Imputing)")

mean(Joindata_t$steps)
median(Joindata_t$steps)
```

After imputing the missing values, the mean and median shift down compared to before.
## Are there differences in activity patterns between weekdays and weekends?

```{r}
# convert date to class 
Joindata$date <- as.Date(Joindata$date)

# add a new column containing day of week
Joindata$weekday = weekdays(Joindata$date)

# factor containing either Weekday OR Weekend
Joindata$weekday_class <- ifelse(Joindata$weekday == "Saturday" | Joindata$weekday == 
                                 "Sunday", "Weekend", "Weekday")
# convert column to factor
Joindata$weekday_class <- factor(Joindata$weekday_class)

#compute median step time for imputing
step_wkd <- aggregate(steps ~ interval + weekday_class, data=Joindata, FUN=mean)


library(lattice)

xyplot(steps ~ interval | weekday_class, data = step_wkd, layout = c(1, 2), type = "l", 
       ylab = "Avg number of steps", 
       main = "Avg number of steps by Day Type")

#The chart shows that the weekdays have one large spike of steps while the weekend has multiple spikes throughout the day.
```
